Lossy compression of DNA sequencing quality data

### 2015-05-02 manuscript/q-compression.tex
Still more revising ...

We compare the \textsc{mse} versus bits/base-call of the
\textit{Rhodobacter sphaeroides}, \textit{Homo sapiens},
\textit{Escherichia coli}, and \textit{Mus musculus} data sets (Figure
\ref{fig:mse_vs_bpbp}). We only include the fragment libraries for the
\textit{Rhodobacter sphaeroides}, and \textit{Homo sapiens} data sets,
but the additional short-jump library results are available in the
Supplemental Table 1. Storing the uncompressed quality values requires
1 byte per base-call because they are stored in \textsc{ascii} format
and is denoted by the dotted black asterisk in the figure. The
lossless compression of each data set using \textsc{bz}ip2 ranges from
2.19 - 3.10 bits/base-call and is denoted by the colored asterisks on
the abscissa. The compression methods tend to cluster together across
the different data sets. Across all data sets, the 0-degree polynomial
regression, profile encodings, and \textsc{q}ual\textsc{c}omp have the
lowest bits/base-call.

\textsc{q}ual\textsc{c}omp with the rate parameter set to 100
bits/read has the lowest \textsc{mse}, but requires 10-15x more
storage than the profile encoding methods for only a 2-3x reduction in
\textsc{mse}. When \textsc{q}ual\textsc{c}omp's rate parameter is set
to match the profile encoding methods, \textsc{q}ual\textsc{c}omp
performs slightly worse in terms of \textsc{mse}. In the
\textit{Rhodobacter sphaeroides} fragment library,
\textsc{q}ual\textsc{c}omp with rate 10 bits/read (0.099 bits/bp) has
a \textsc{mse} of 17.29. Using 256-profile encoding requires less
storage (0.079 bits/bp) and has a lower \textsc{mse} (11.85).

@@

@@

Preprocessing involves two steps: \emph{trimming} the poor quality
regions of the sequences and \emph{discarding} sequences that are
deemed to be poor quality overall. After trimming, the majority of
compression methods retain more base-pairs than the uncompressed
sequences (Figure \ref{fig:preprocessing}). In general, as a given
compression model increases in complexity, i.e., as the number of
profiles, polynomial degrees, or rate increases, the amount of
base-pairs kept approaches the number base-pairs kept using the
uncompressed sequences. The compression methods on the \textit{Mus
  musculus} data set have the greatest proportion of retained
base-pairs compared to the uncompressed sequences. The
\textit{Escherichia coli} str. K-12 MG1655 MiSeq data set has the
smallest range.

The 2-bin approach is the only compression method that results in a
higher number of trimmed base-pairs across all data sets. Sickle uses
a sliding window approach to smooth the read quality values before it
trims. In the 2-bin approach, there is not an even distribution of
values per bin. In other words, \emph{bad} quality values may range
from 0-33, whereas \emph{good} values may only range from 34-40. Thus,
mid-range quality values that are above the threshold (20 by default)
are set below the quality threshold when compressed, resulting in an
increased number of trimmed bases.

@@

Perhaps one of the greatest potential benefits of compressing quality
values is the potential to perform quality control and possibly other
operations directly on the compressed representations of the
data. This is easiest to to consider for profile-based
compression. The $k$ profiles can be evaluated for (pre-)processing
operations such as filtering and trimming, and the operations
transitively applied to the entire set of reads, thus saving
substantial computation associated with evaluating the full set of
reads.

@@

In this paper we have examined lossy compression on sequence quality
values and their effect on subsequent analyses. Although most previous
examinations on lossy compression primarily focused on information
loss, we have shown that typically used bioinformatics software today
have additional built-in sensitivity to handle significant loss of
information in the compressed quality values.

>>>>

We compare the \textsc{mse} versus bits/base-call of the
\textit{Rhodobacter sphaeroides}, \textit{Homo sapiens},
\textit{Escherichia coli}, and \textit{Mus musculus} data sets (Figure
\ref{fig:mse_vs_bpbp}). We only include the fragment libraries for the
\textit{Rhodobacter sphaeroides}, and \textit{Homo sapiens} data sets,
but the additional short-jump library results are available in the
Supplemental Table 1. Storing the uncompressed quality values requires
1 byte per base-call because they are stored in \textsc{ascii} format
and is denoted by the dotted black asterisk in the figure. The
lossless compression of each data set using \textsc{bz}ip2 ranges from
2.19--3.10 bits/base-call and is denoted by the colored asterisks on
the abscissa. The compression methods tend to cluster together across
the different data sets. Across all data sets, the 0-degree polynomial
regression, profile encodings, and \textsc{q}ual\textsc{c}omp have the
lowest bits/base-call.

\textsc{q}ual\textsc{c}omp with the rate parameter set to 100
bits/read has the lowest \textsc{mse}, but requires
$\sim$10--15$\times$ more storage than the profile encoding methods
for only a $\sim$2--3$\times$ reduction in \textsc{mse}. When
\textsc{q}ual\textsc{c}omp's rate parameter is set to match the
profile encoding methods, \textsc{q}ual\textsc{c}omp performs slightly
worse in terms of \textsc{mse}. In the \textit{Rhodobacter
  sphaeroides} fragment library, \textsc{q}ual\textsc{c}omp with rate
10 bits/read (0.099 bits/bp) has a \textsc{mse} of 17.29. Using
256-profile encoding requires less storage (0.079 bits/bp) and has a
lower \textsc{mse} (11.85).

@@

As an additional point of reference regarding compression
effectiveness the recently published
Read-Quality-Sparsifier~\citep{DBLP:conf/recomb/YuYB14} achieved best
case compression of 0.254 bits/base-call, and a mean compression of
1.841 bits/base-call; both values higher than all of the profile
results and for some results of other methods presented here
(Figure~\ref{fig:mse_vs_bpbp}).

@@

Preprocessing involves two steps: \emph{trimming} the poor quality
regions of the sequences and \emph{discarding} sequences that are
deemed to be poor quality overall. After trimming, the majority of
compression methods retain more base-pairs than the uncompressed
sequences (Figure \ref{fig:preprocessing}). In general, as a given
compression model increases in complexity, i.e., as the number of
profiles, polynomial degrees, or rate increases, the amount of
base-pairs kept approaches the number base-pairs kept using the
uncompressed sequences. The compression methods on the \textit{Mus
  musculus} data set have the greatest proportion of retained
base-pairs compared to the uncompressed sequences. The
\textit{Escherichia coli} MiSeq data set has the smallest range.

The 2-bin approach is the only compression method that results in a
higher number of trimmed base-pairs across all data sets. Sickle uses
a sliding window approach to smooth the read quality values before it
trims. In the 2-bin approach, there is not an even distribution of
values per bin. In other words, \emph{bad} quality values may range
from 0--33, whereas \emph{good} values may only range from
34--40. Thus, mid-range quality values that are above the threshold
(20 by default) are set below the quality threshold when compressed,
resulting in an increased number of trimmed bases.

@@

Perhaps one of the greatest potential benefits of compressing quality
values is the ability to perform quality control and possibly other
operations directly on the compressed representations of the data. For
example, with profile-based compression the $k$ profiles can be
evaluated for (pre-)processing operations such as filtering and
trimming, and the operations transitively applied to the entire set of
reads, thus saving substantial computation.

@@

In this paper we have examined lossy compression on sequence quality
values and their effect on subsequent analyses. Although most previous
examinations on lossy compression focused primarily on information
loss and compression effectiveness, we show introduce several general
approaches to quality value compression and show that common
bioinformatic analyses are robust substantial loss of information in
quality values.
